---
title: "covbat_harmonization_cognition_ses"
output: html_document
date: "2025-09-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# https://github.com/andy1764/ComBatFamily
# https://github.com/andy1764/CovBat_Harmonization/blob/master/R/R/covbat.R
install.packages("remotes")
remotes::install_github("andy1764/ComBatFamily")
install.packages("devtools")
devtools::install_github("andy1764/CovBat_Harmonization/R")
```

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(splines)
  library(ComBatFamQC)
  library(limma)
})
```

# 0) load data
outputs from `HBN_replication_regression.ipynb` and `Step1_Data_Prep.ipynb`, respectively
```{r}
# ROOT_DIR <- "/Users/bmacedo/Desktop/final_WM"
ROOT_DIR = "/mnt/isilon/bgdlab_hbcd/projects/macedo_wm_exposome/macedo_wm_exposome"

# Directories
CLEANED_DIR <- file.path(ROOT_DIR, "output_data", "cleaned")
HBN_DIR <- file.path(CLEANED_DIR, "HBN")

# Load data
hbn_raw  <- read.csv(file.path(HBN_DIR, "final_hbn_dataset_12_23.csv"))
abcd_raw <- read.csv(file.path(CLEANED_DIR, "exposome_FINAL_cognition_11_10.csv"))
```

# 1) standardize key covariate names to match across datasets
```{r}
hbn_df <- hbn_raw %>% rename(age = Age, sex = Sex, qc_prediction = t1_neighbor_corr)
```


# 2) rename HBN measurement columns to match abcd syntax
```{r}
feature_cols <- grep("^FEATURES_", names(hbn_df), value = TRUE)

make_bundle_name <- function(col) {
  rest <- sub("^FEATURES_", "", col)
  parts <- strsplit(rest, "_", fixed = TRUE)[[1]]

  if (length(parts) < 3) return(col)

  bundle_id <- paste(parts[1:2], collapse = "_")
  metric <- paste(parts[3:length(parts)], collapse = "_")

  paste0("bundle_", bundle_id, "_", metric)
}

new_names <- vapply(feature_cols, make_bundle_name, character(1))

# apply rename only to those columns
names(hbn_df)[match(feature_cols, names(hbn_df))] <- new_names

```

# 3) add dataset label, create unique batch IDs
```{r}
hbn_df  <- hbn_df  %>% mutate(dataset = "HBN",  batch = paste0("HBN_",  site))
abcd_df <- abcd_raw %>% mutate(dataset = "ABCD", batch = paste0("ABCD_", site))

ref_batch <- "ABCD_16"  # anchor
```

# 4) choose harmonization feature columns
```{r}
hbn_msmt_cols  <- grep("^bundle_", names(hbn_df),  value = TRUE)
abcd_msmt_cols <- grep("^bundle_", names(abcd_df), value = TRUE)
msmt_cols <- intersect(hbn_msmt_cols, abcd_msmt_cols)

cat("HBN msmt cols:", length(hbn_msmt_cols), "\n")
cat("ABCD msmt cols:", length(abcd_msmt_cols), "\n")
cat("Intersect msmt_cols:", length(msmt_cols), "\n")
```

```{r}
get_bundle <- function(x) sub("^bundle_([^_]+_[^_]+)_.*$", "\\1", x)

hbn_bundles  <- sort(unique(get_bundle(hbn_msmt_cols)))
abcd_bundles <- sort(unique(get_bundle(abcd_msmt_cols)))

cat("Bundles in ABCD not HBN:\n"); print(setdiff(abcd_bundles, hbn_bundles))
cat("\nBundles in HBN not ABCD:\n"); print(setdiff(hbn_bundles, abcd_bundles))

```

# 5) combine datasets for harmonization
```{r}
hbn_df$site <- as.character(hbn_df$site)
abcd_df$site <- as.character(abcd_df$site)

all_df <- bind_rows(hbn_df, abcd_df)
all_df$sex <- as.factor(all_df$sex)
```

# 6) Build input matrices for harmonization
#    dat: features x subjects
#    mod: covariates to protect
```{r}
dat <- t(as.matrix(all_df[, msmt_cols]))  # (features x subjects)
mod <- model.matrix(~ ns(age, df = 3) + sex + General_SES, data = all_df)
```

# 7) harmonize, anchored to ABCD site 16
```{r}
all_df$batch <- factor(trimws(all_df$batch))
ref_batch <- trimws(as.character(ref_batch))

harm <- covfam(data=as.matrix(all_df[, msmt_cols]), 
               bat=all_df$batch, 
               covar=all_df[, c("age","sex")], 
               model=mgcv::gam, 
               formula=y ~ s(age, k=3) + sex + General_SES, 
               eb=TRUE, 
               ref.batch=ref_batch)

X_harm <- harm$dat.covbat
all_df_harm <- all_df
all_df_harm[, msmt_cols] <- X_harm
```

# 8) Keep only HBN rows, rebuild output df with harmonized features
```{r}
HARMONIZED_DIR <- file.path(ROOT_DIR, "output_data", "harmonized", "HBN")
if (!dir.exists(HARMONIZED_DIR)) {dir.create(HARMONIZED_DIR, recursive = TRUE)}

hbn_out <- all_df_harm[all_df_harm$dataset == "HBN", , drop = FALSE] %>% as.data.frame()

# Define which non-msmt columns you want to keep
keep_meta <- c("participant_id", "site", "batch", "dataset", "age", "sex", "qc_prediction", "General_SES", "matched_group")

# Keep ONLY harmonized features + metadata
hbn_out <- hbn_out[, c(keep_meta, msmt_cols), drop = FALSE]

# save
output_path <- file.path(HARMONIZED_DIR, "HBN_harmonized_refABCD16_01_20.csv")
write.csv(hbn_out, file = output_path, row.names = FALSE)
cat("[SAVE] →", output_path, "\n")
```

# 9) Nuisance regression on HBN data
```{r}
# make sure sex is a factor and drop unused levels
hbn_out$sex <- droplevels(as.factor(hbn_out$sex))

# design matrix: samples x covariates
# note that qc_prediction == t1_neighborhood_corr here 
Z <- model.matrix(~ ns(age, df = 3) + sex + qc_prediction, data = hbn_out)

# geatures matrix: features x samples
Y <- t(as.matrix(hbn_out[, msmt_cols]))  # (features x samples)'

# fit all features at once
fit <- lmFit(Y, Z)

#  predicted nuisance component (features x samples)
Y_hat <- fit$coefficients %*% t(Z)

# residuals (features x samples), then write back as samples x features
Y_resid <- Y - Y_hat
hbn_out[, msmt_cols] <- t(Y_resid)

```

# 10) Save output
```{r}
# save this version for multivariate replication (after residualizing)
output_path <- file.path(HARMONIZED_DIR, "HBN_harmonized_refABCD16_then_nuisanceResid_12_23.csv")

write.csv(hbn_out, file = output_path, row.names = FALSE)

cat("[SAVE] →", output_path, "\n")
```
















