s---
title: "covbat_harmonization_cognition_ses"
output: html_document
date: "2025-09-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
options(repos = c(CRAN="https://cloud.r-project.org"))
install.packages("remotes")
remotes::install_github("andy1764/ComBatFamily")
remotes::install_github("andy1764/CovBat_Harmonization", dependencies=TRUE)
```

```{r}
# https://github.com/andy1764/ComBatFamily
# https://github.com/andy1764/CovBat_Harmonization/blob/master/R/R/covbat.R
install.packages("remotes")
remotes::install_github("andy1764/ComBatFamily")
install.packages("devtools")
devtools::install_github("andy1764/CovBat_Harmonization/R")
```

```{r}
library(ComBatFamily)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
# which analysis are we running? set that analysis to TRUE and all others to FALSE
main_analysis = TRUE
cognition_analysis = FALSE
income_analysis = FALSE

# assserts that only one analysis is selected
if(sum(main_analysis+cognition_analysis+income_analysis) != 1) {
  stop("You must select exactly one analysis to run")
}
```

Load in data, outputs from `Step1_Data_Prep.ipynb` 
```{r}
# Define root directory
# ROOT_DIR <- "/Users/bmacedo/Desktop/final_WM"
ROOT_DIR <- "/mnt/isilon/bgdlab_hbcd/projects/macedo_wm_exposome/macedo_wm_exposome"

# Define subdirectories
OUTPUT_DIR <- file.path(ROOT_DIR, "output_data")
CLEANED_DIR <- file.path(OUTPUT_DIR, "cleaned")
HARMONIZED_DIR <- file.path(OUTPUT_DIR, "harmonized")

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(CLEANED_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(HARMONIZED_DIR, recursive = TRUE, showWarnings = FALSE)

ref_batch <- "16"

if (main_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "exposome_FINAL_11_3.csv"))
}

if (cognition_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "exposome_FINAL_cognition_11_10.csv"))
}

if (income_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "parental_edu_income_sensitivity_df.csv"))
}
```

Create vectors of training and testing data for harmonization
We only use the training data for harmonization procedure to avoid data leakage
```{r}
# create a logical vector for covbat function so it knows which data to use for harmonization
# when A is training data
exposome_group_vector_A_train = ifelse(exposome_df$matched_group == 1, TRUE, FALSE)
# when B is training data
exposome_group_vector_B_train = ifelse(exposome_df$matched_group == 2, TRUE, FALSE) 
n_A <- sum(exposome_group_vector_A_train)
n_B <- sum(exposome_group_vector_B_train)

cat("Group A:", n_A, "\nGroup B:", n_B, "\n")
```
main_analysis: Group A: 4078, Group B: 4097s 
cognition_analysis: Group A: 3921, Group B: 3956
edu_sensitivity: Group A: 3371, Group B: 3438

```{r}
exposome_batch = as.factor(exposome_df$site) # batch is the site
exposome_batch <- relevel(exposome_batch, ref = "16")

# optional check
table(exposome_batch, useNA = "ifany")
levels(exposome_batch)
```

```{r}
# extract  the features that we want to harmonize (all metrics for all bundles)
# select only columns that start with "msmt_"
msmt_cols <- grep("^bundle_", names(exposome_df), value = TRUE)

# subset dataframe to only those columns
data_exposome_clean <- exposome_df[, msmt_cols]

# convert to transposed numeric matrix for input into covbat
data_exposome <- data.matrix(data_exposome_clean)
```

Select which demographic covariates we want CovBat to protect: signal from these covariates will be maintained even after harmonization
```{r}
# we protect age, sex, head motion, and exposome scores

# build covariate dataframe and linear formula
if (main_analysis) {
  cov_exposome <- exposome_df[, c(
    "age", "sex",
    "General_SES", "School", "Family_Values", "Family_Turmoil",
    "Dense_Urban_Poverty", "Extracurriculars", "Screen_Time"
  )]

  lm_formula <- y ~ age + sex +
    General_SES + School + Family_Values + Family_Turmoil +
    Dense_Urban_Poverty + Extracurriculars + Screen_Time
}

if (cognition_analysis) {
  cov_exposome <- exposome_df[, c(
    "age", "sex",
    "General_SES", "School", "Family_Values", "Family_Turmoil",
    "Dense_Urban_Poverty", "Extracurriculars", "Screen_Time",
    "neurocog_pc1.bl", "neurocog_pc2.bl", "neurocog_pc3.bl"
  )]

  lm_formula <- y ~ age + sex +
    General_SES + School + Family_Values + Family_Turmoil +
    Dense_Urban_Poverty + Extracurriculars + Screen_Time +
    neurocog_pc1.bl + neurocog_pc2.bl + neurocog_pc3.bl
}

if (income_analysis) {
  cov_exposome <- exposome_df[, c(
    "age", "sex", "qc_prediction", "parental_education", "income"
  )]

  lm_formula <- y ~ age + sex + qc_prediction + parental_education + income
}
```

Run the covbat algorithm 
```{r}
# -------------------------
# A as training set
# -------------------------
train_idx_A <- exposome_group_vector_A_train
test_idx_A  <- !train_idx_A

# optional safety check:
if (!any(exposome_batch[train_idx_A] == ref_batch)) {
  stop("Reference site 16 is not present in A training data")
}

covbat_exposome_A_fit <- covfam(
  data      = data_exposome[train_idx_A, , drop = FALSE],
  bat       = droplevels(exposome_batch[train_idx_A]),
  covar     = cov_exposome[train_idx_A, , drop = FALSE],
  model     = lm,
  formula   = lm_formula
)

# harmonized training data
covbat_exposome_A_train_harmonized <- covbat_exposome_A_fit$dat.covbat

# apply learned harmonization parameters to test data
covbat_exposome_A_test_pred <- predict(
  covbat_exposome_A_fit,
  newdata  = data_exposome[test_idx_A, , drop = FALSE],
  newbat   = factor(as.character(exposome_batch[test_idx_A]),
                    levels = levels(droplevels(exposome_batch[train_idx_A]))),
  newcovar = cov_exposome[test_idx_A, , drop = FALSE]
)

covbat_exposome_A_test_harmonized <- covbat_exposome_A_test_pred$dat.covbat

# -------------------------
# B as training set
# -------------------------
train_idx_B <- exposome_group_vector_B_train
test_idx_B  <- !train_idx_B

# optional safety check:
if (!any(exposome_batch[train_idx_B] == ref_batch)) {
  stop("Reference site 16 is not present in B training data")
}

covbat_exposome_B_fit <- covfam(
  data      = data_exposome[train_idx_B, , drop = FALSE],
  bat       = droplevels(exposome_batch[train_idx_B]),
  covar     = cov_exposome[train_idx_B, , drop = FALSE],
  model     = lm,
  formula   = lm_formula,
  ref.batch = ref_batch
)

# harmonized training data
covbat_exposome_B_train_harmonized <- covbat_exposome_B_fit$dat.covbat

# apply learned harmonization parameters to test data
covbat_exposome_B_test_pred <- predict(
  covbat_exposome_B_fit,
  newdata  = data_exposome[test_idx_B, , drop = FALSE],
  newbat   = factor(as.character(exposome_batch[test_idx_B]),
                    levels = levels(droplevels(exposome_batch[train_idx_B]))),
  newcovar = cov_exposome[test_idx_B, , drop = FALSE]
)

covbat_exposome_B_test_harmonized <- covbat_exposome_B_test_pred$dat.covbat

```

```{r}
if (!dir.exists(HARMONIZED_DIR)) { dir.create(HARMONIZED_DIR, recursive = TRUE) }

## extract harmonized metrics
harmonized_exposome_A <- t(covbat_exposome_A$dat.covbat)
harmonized_exposome_B <- t(covbat_exposome_B$dat.covbat)

## combine harmonized metrics back with our demographic and exposome measures
msmt_idx <- grep("^msmt_", names(exposome_df))
df_harmonized_exposome_A <- cbind(exposome_df[, -c(msmt_idx)], harmonized_exposome_A)
df_harmonized_exposome_B <- cbind(exposome_df[, -c(msmt_idx)], harmonized_exposome_B)

## define filenames
if (main_analysis) {
  fileA <- "df_harmonized_exposome_A_12_10.csv"
  fileB <- "df_harmonized_exposome_B_12_10.csv"
}

if (cognition_analysis) {
  fileA <- "df_harmonized_exposome_cognition_A_12_10.csv"
  fileB <- "df_harmonized_exposome_cognition_B_12_10.csv"
}

if (income_analysis) {
  fileA <- "df_income_sens_A_11_30.csv"
  fileB <- "df_income_sens_B_11_30.csv"
}

## full paths
outputfileA <- file.path(HARMONIZED_DIR, fileA)
outputfileB <- file.path(HARMONIZED_DIR, fileB)

## save
write.csv(df_harmonized_exposome_A, file = outputfileA, row.names = FALSE)
write.csv(df_harmonized_exposome_B, file = outputfileB, row.names = FALSE)

cat("[SAVE] A →", outputfileA, "\n")
cat("[SAVE] B →", outputfileB, "\n")
```


















