---
title: "covbat_harmonization_cognition_ses"
output: html_document
date: "2025-09-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# https://github.com/andy1764/ComBatFamily
# https://github.com/andy1764/CovBat_Harmonization/blob/master/R/R/covbat.R
install.packages("remotes")
remotes::install_github("andy1764/ComBatFamily")
install.packages("devtools")
devtools::install_github("andy1764/CovBat_Harmonization/R")
```

```{r}
library(ComBatFamily)
library(CovBat)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
# which analysis are we running? set that analysis to TRUE and all others to FALSE
main_analysis = FALSE
cognition_analysis = FALSE
income_analysis = TRUE

# assserts that only one analysis is selected
if(sum(main_analysis+cognition_analysis+income_analysis) != 1) {
  stop("You must select exactly one analysis to run")
}
```

Load in data, outputs from `Step1_Data_Prep.ipynb` 
```{r}
# Define root directory
ROOT_DIR <- "/Users/bmacedo/Desktop/final_WM"

# Define subdirectories
OUTPUT_DIR <- file.path(ROOT_DIR, "output_data")
CLEANED_DIR <- file.path(OUTPUT_DIR, "cleaned")
HARMONIZED_DIR <- file.path(OUTPUT_DIR, "harmonized")

if (main_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "exposome_FINAL_11_3.csv"))
}

if (cognition_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "exposome_FINAL_cognition_11_10.csv"))
}

if (income_analysis) {
  exposome_df <- read.csv(file.path(CLEANED_DIR, "parental_edu_income_sensitivity_df.csv"))
}
```

Create vectors of training and testing data for harmonization
We only use the training data for harmonization procedure to avoid data leakage
```{r}
# create a logical vector for covbat function so it knows which data to use for harmonization
# when A is training data
exposome_group_vector_A_train = ifelse(exposome_df$matched_group == 1, TRUE, FALSE)
# when B is training data
exposome_group_vector_B_train = ifelse(exposome_df$matched_group == 2, TRUE, FALSE) 

n_A <- sum(exposome_group_vector_A_train)
n_B <- sum(exposome_group_vector_B_train)

cat("Group A:", n_A, "\nGroup B:", n_B, "\n")
```
main_analysis: Group A: 4185, Group B: 4243 
cognition_analysis: Group A: 3921, Group B: 3956
edu_sensitivity: Group A: 3371, Group B: 3438

```{r}
exposome_batch = as.factor(exposome_df$site) # batch is the site
```

```{r}
# extract  the features that we want to harmonize (all metrics for all bundles)
# select only columns that start with "msmt_"
msmt_cols <- grep("^msmt_", names(exposome_df), value = TRUE)

# subset dataframe to only those columns
data_exposome_clean <- exposome_df[, msmt_cols]

# convert to transposed numeric matrix for input into covbat
data_exposome <- data.matrix(data_exposome_clean)
data_exposome <- t(data_exposome) # (features x subjects)
```

Select which demographic covariates we want CovBat to protect: signal from these covariates will be maintained even after harmonization
```{r}
# we protect age, sex, head motion, and exposome scores

if(main_analysis) {
  mod_exposome <- model.matrix(~ age + sex +
                               General_SES + School + Family_Values + Family_Turmoil +
                               Dense_Urban_Poverty + Extracurriculars + Screen_Time,
                             data = exposome_df)}

if(cognition_analysis){
  mod_exposome <- model.matrix(~ age + sex + 
                               General_SES + School + Family_Values + Family_Turmoil +
                               Dense_Urban_Poverty + Extracurriculars + Screen_Time +
                               neurocog_pc1.bl + neurocog_pc2.bl + neurocog_pc3.bl,
                             data = exposome_df)}

if(income_analysis){
  mod_exposome <- model.matrix(~ age + sex + qc_prediction + parental_education + income,
                             data = exposome_df)}

```

Run the covbat algorithm 
```{r}
covbat_exposome_A <- covbat(dat = data_exposome,
                        bat = exposome_batch,
                        mod = mod_exposome,
                        train = exposome_group_vector_A_train)

covbat_exposome_B <- covbat(dat = data_exposome,
                        bat = exposome_batch,
                        mod = mod_exposome,
                        train = exposome_group_vector_B_train)

```

```{r}
if (!dir.exists(HARMONIZED_DIR)) { dir.create(HARMONIZED_DIR, recursive = TRUE) }

## extract harmonized metrics
harmonized_exposome_A <- t(covbat_exposome_A$dat.covbat)
harmonized_exposome_B <- t(covbat_exposome_B$dat.covbat)

## combine harmonized metrics back with our demographic and exposome measures
msmt_idx <- grep("^msmt_", names(exposome_df))
df_harmonized_exposome_A <- cbind(exposome_df[, -c(msmt_idx)], harmonized_exposome_A)
df_harmonized_exposome_B <- cbind(exposome_df[, -c(msmt_idx)], harmonized_exposome_B)

## define filenames
if (main_analysis) {
  fileA <- "df_harmonized_exposome_A_12_10.csv"
  fileB <- "df_harmonized_exposome_B_12_10.csv"
}

if (cognition_analysis) {
  fileA <- "df_harmonized_exposome_cognition_A_12_10.csv"
  fileB <- "df_harmonized_exposome_cognition_B_12_10.csv"
}

if (income_analysis) {
  fileA <- "df_income_sens_A_11_30.csv"
  fileB <- "df_income_sens_B_11_30.csv"
}

## full paths
outputfileA <- file.path(HARMONIZED_DIR, fileA)
outputfileB <- file.path(HARMONIZED_DIR, fileB)

## save
write.csv(df_harmonized_exposome_A, file = outputfileA, row.names = FALSE)
write.csv(df_harmonized_exposome_B, file = outputfileB, row.names = FALSE)

cat("[SAVE] A →", outputfileA, "\n")
cat("[SAVE] B →", outputfileB, "\n")
```


















